\chapter{Introduction}

\section{Background}

It is generally useful for software to be able to programmatically determine the
capabilities of the hardware it is running on. In addition external debuggers
need to know the same information. Instead of having each RISC-V extension
specify their own discovery mechanism, this document attempts to specify a
single format and access method that can be used to describe any feature
software and debuggers might want to be aware of.

\section{Goals}

\begin{steps}{The task group will define:}
\item A machine-readable format for system descriptions. It's intended to be
present in hardware when power on, so software and debuggers can read this
description. 
\item A mechanism for software and debuggers to access the system description,
if it exists.
\item A human-readable format for system descriptions. It's intended to be used
to talk about the system description in this document as well as other
documentation, to help designers write a system description, and for debuggers
to display the system description to their users.
\item Software that translates back and forth between the machine-readable and
human-readable format.
\item Defining the equivalent syntax and format of system description in external 
industrial standards. Software translates system description into the corresponding 
definitions according to the external specifications for system boot process later 
on.

\begin{commentary}
The external industrial standards refer to Device Tree, SMBIOS, ACPI and etc. 
specifications which are defined for different computer segments. These industrial 
specifications are typically used to describe hardware features and configurations.
\end{commentary}

\end{steps}

\section{Use Cases}

\subsection{External Debuggers}

When an external debugger connects to a system, it would like to discover as
much as possible about that system in as little time as possible. Some of this
is merely to show the user (e.g. a manufacturer name), while other features are
critical to the user (e.g. XLEN), and other features determine what kind of
operations the user can perform (e.g. supported hardware trigger types). Most
of these are already discoverable, although many require writing a value and
checking the result to see whether support exists.

In the case of hardware trigger types there are so many possible combinations
of options that this is prohibitive. In other cases the software to detect all
features is quite complex. When some harts are powered down, no discovery is
possible on them at all. Supplying a system description can solve all these
problems.

\subsection{System Firmware}
Supplying system descriptions to system firmware for discovering processor 
core features and platform configurations. Based on system descriptions, firmware 
can determine the hardware capabilities in programmatic access method 
(Section~\ref{sec:AccessMethod}) and configure hardware accordingly. This instead 
of testing on specific hardware feature and handling the fail condition or customize 
system firmware in order to compliant with particular hardware features.

Most of the use cases are determination of availability of optional hardware 
features such as physical memory protection (PMP) unit, enhanced features of PMP,
the availability of Core Local Interrupt Controller (CLIC), Core Local interruptor 
(CLINT) memory map any future optional core features. The settings of Platform-Level 
Interrupt Controller (PLIC), availability of platform Real time Clock (RTC),
Reset mechanism and etc. are also in the scope of system descriptions for platforms.

System firmware may convert system descriptions into other representations which are 
defined in the external industrial standards (Appendix~\ref{sec:ExternalIndustrialStandard})
and passed to the subsequent boot processes.

\begin{commentary}
The system firmware stated herein refer to the startup software which is executed 
when system power on. The system firmware initializes hardware configuration and 
builds up firmware services or data structures for booting up system to OS. The typical
system firmware such as uboot for embedded systems, BIOS for PCs or other firmware 
frameworks.

\end{commentary}

\chapter{Machine-Readable Format}

\begin{steps}{Ideas:}
\item Fully custom, along the lines of
    \url{https://github.com/riscv/riscv-debug-spec/pull/450/files#diff-9cee87c1f0feb10d9e3e9b2bfad92985R380}
\item Semi-custom, using CBOR
\item ...
\end{steps}

\chapter{Human-Readable Format}

\begin{steps}{Ideas:}
\item Take the binary format, and expand constants to strings, add whitespace,
keeping the mapping as close to 1:1 as possible.
\item XML
\item JSON
\item ...
\end{steps}

\chapter{Access Method}
\label{sec:AccessMethod}

\begin{steps}{Ideas:}
\item A CSR contains the base address of where the structure can be accessed on
the system bus
\item A CSR address/data pair can be used to access the structure. (Write index
to the address CSR, then read from the data CSR.)
\item ...
\end{steps}
